---
layout: post
title: "Downloading, cleaning and analyzing, WDI indicators in R with codes and examples"
date: 2020-11-02
categories: R
# status: process
# published: true
status: development
published: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	fig.align = "center",
	fig.height = 6,
	fig.width = 8,
	message = FALSE,
	warning = FALSE
)
```

# **What’s in this blog?** 

This blog is about performing all the following tasks using [R programming language](https://www.r-project.org/about.html) with expamples, codes and step by step explanation. 

- How to search your variable(s) of interest, in the [World Bank](http://www.worldbank.org/), from [International Debt Statistics](https://data.worldbank.org/products/ids)  to the [World Development Indicators (WDI)](http://datatopics.worldbank.org/world-development-indicators/)?

- How to download data for the variable(s), ranging from 1960 to the latest?

- How to clean the data and readying it for your analysis?

- How to visualize your statistic? 


# **Example for explaining the above tasks** 

## **Variables/Indicators**

1. Military expenditure (% of GDP) 

2. Government expenditure on education, total (% of GDP)

3. Current health expenditure (% of GDP)

## **Countries**

1. Bangladesh.

2. India 

3. Pakistan 


# **1. Getting Data on required Indicators**

First, we need to search indicators and get codes for these. I am searching these one by one. You may do these at once.

## 1.1 Searching indicators and get codes

### 1.1.1 Search for military expenditure

```{r echo=TRUE, fig.align='center', fig.height=6, fig.width=12, warning=FALSE, dpi=400}
library(WDI)

WDIsearch(string="military expenditure",
          short = TRUE)
```

From the first search result I found ***Military expenditure (% of GDP)*** and its code ***MS.MIL.XPND.GD.ZS***

### 1.1.2 Search for health expenditure

```{r echo=TRUE,fig.align='center', fig.height=6, fig.width=12, warning=FALSE, dpi=400}

WDIsearch(string="current health expenditure",
          short = TRUE)
```

From the second search result I found ***Current health expenditure (% of GDP)*** and its code ***SH.XPD.CHEX.GD.ZS***

### 1.1.3 Search for expenditure on education

```{r echo=TRUE, fig.align='center', fig.height=6, fig.width=12, warning=FALSE, dpi=400}

WDIsearch(string="government expenditure on education",
          short = TRUE)
```

From the third search result I found ***Government expenditure on education, total (% of GDP)*** and its code ***SE.XPD.TOTL.GD.ZS***

## 1.2 Getting data for selected indicators

```{r echo=TRUE, fig.align='center', fig.height=6, fig.width=12, warning=FALSE, dpi=400}

DHEexp_data<- WDI(country = c("BGD","IND","PAK"), 
    indicator = c('Military_expenditure' = 'MS.MIL.XPND.GD.ZS',
                  'Health_expenditure' = 'SH.XPD.CHEX.GD.ZS',
                  'Education_expenditure' = 'SE.XPD.TOTL.GD.ZS'),
    start=1960, end=2020)

```

Now let's have a look at downloaded data

```{r echo=TRUE, fig.align='center', fig.height=6, fig.width=12, warning=FALSE, dpi=400}

library(tibble)

as_tibble(DHEexp_data)

```


# **2. Cleaning the Data**

As I need years for which data is available. Therefore, observations with NAs can be removed like this.  

```{r echo=TRUE, fig.align='center', fig.height=6, fig.width=12, warning=FALSE, dpi=400}

library(tidyr)
library(dplyr)


DHEexp_data <- drop_na(DHEexp_data)

as_tibble(DHEexp_data)

```

As there are three countries, so lets filter years which are repeated 3 times, then join it with original data. While remaining are removed. Seven observations are for each country now. 

```{r echo=TRUE, fig.align='center', fig.height=6, fig.width=12, warning=FALSE, dpi=400}

yr3 <- count(DHEexp_data,year, sort = TRUE)

DHEexp_data <-left_join(DHEexp_data,yr3, by="year") 

DHEexp_data <- filter(DHEexp_data, n==3)

DHEexp_data <- select(DHEexp_data,1:6)

as_tibble(DHEexp_data)

```

To get the final data for visualization, data is reshaped. Subsequently rounded and selected only three years: 

- $2000$ 

- $2006$ 

- $2012$

```{r echo=TRUE, fig.align='center', fig.height=6, fig.width=12, warning=FALSE, dpi=400}

DHEexp_data<- pivot_longer(DHEexp_data,
                           !c("iso2c","country","year"), 
                           names_to = "expenditures",
                           values_to = "values")

DHEexp_data <- DHEexp_data %>% mutate_if(is.numeric,
                                         round, digits=2)

plotdata <- filter(DHEexp_data,year==2000 |
                     year==2006|
                     year==2012)

as_tibble(plotdata)

```


# **3. Data visualizations**


```{r echo=FALSE, fig.align='center', fig.height=6, fig.width=12, warning=FALSE, dpi=400}

library(ggplot2)

mytheme<- theme(plot.caption = element_text(size = 12, face = "bold"),
                legend.title = element_blank(),
                legend.key.size = unit(0.05, "npc"),
                legend.position = "top",
                plot.title = element_text(size = 14, hjust = 0.5, 
                                          face = "bold"),
                legend.text = element_text(size = 12),
                axis.title = element_text(size = 12, face = "bold"),
                axis.text.x = element_text(size = 12, face = "bold"),
                axis.text.y = element_text(size = 12, face = "bold"),
                plot.caption.position = "plot",
                strip.text = element_text(size = 12, face = "bold")
)

myplot <- aes(x =country, y = values,
              fill = factor(year))

mygeom <- geom_col(position = "dodge",width = 0.5)

myguides<- guides(fill=guide_legend(nrow = 1))

mycaption <- "By: Ijaz Ali,IBA Karachi" 

```

## **3.1 Military expenditure (% of GDP)**


```{r echo=TRUE, fig.align='center', fig.height=6, fig.width=12, warning=FALSE, dpi=400}

ggplot(data = filter(plotdata,expenditures=="Military_expenditure"), 
       mapping = myplot) + mygeom + mytheme + myguides +
  labs(x="", y="",
       caption = mycaption,
       title = "Military expenditure (% of GDP)")

```



## **3.2 Current health expenditure (% of GDP)**


```{r echo=TRUE, fig.align='center', fig.height=6, fig.width=12, warning=FALSE, dpi=400}

ggplot(data = filter(plotdata,expenditures=="Health_expenditure"), 
        mapping = myplot) + mygeom + mytheme + myguides +
        labs(x="", y="",
        caption = mycaption,
        title = "Current health expenditure (% of GDP)")

```


## **3.3 Government expenditure on education, total (% of GDP)**


```{r echo=TRUE, fig.align='center', fig.height=6, fig.width=12, warning=FALSE, dpi=400}

ggplot(data = filter(plotdata,expenditures=="Education_expenditure"), 
       mapping = myplot) + mygeom + mytheme + myguides +
      labs(x="", y="",
       caption = mycaption,
       title = "Government expenditure on education, total (% of GDP)")

```









